\begin{center}
	\textbf{Budget and Budget Justification}
\end{center}

\noindent I am requesting \$25,000 in funding from the AI-Informed Seed Grant program. These funds will primarily be spent on labor costs for two collaborators, whom I will direct and supervise:
\begin{itemize}
	\item Matthew N.\ White (JHU PhD 2014), a full time employee of Econ-ARK who was the primary developer for the \texttt{HARK} software package and has collaborated with me on heterogeneous agents macroeconomic research projects. 
	
	\item Marc Maliar (UChicago BA 2022), an aspiring economist who served as a programmer for the TensorFlow code used for \cite{MALIAR202176} and has intimate knowledge of its development. 
\end{itemize}

\noindent Marc's primary task is simply to ``onboard'' White and me, so that we are familiar with the MMW TensorFlow code structure and can reproduce its results locally. We already have our own code for a benchmark version of the Krusell-Smith model, providing an additional source of verification. I will then work with Dr.\ White to more fully characterize the nature of the differences between solutions produced by the methods.

\vspace{0.2cm}

\noindent Conservatively budgeting \$3000 of Marc Maliar's time to bring us up to speed (and for additional consulting later), and providing my own research time gratis, this leaves over six weeks of Dr.\ White's full time. Given the well bounded nature of this seed project, our ``hot start'' on the work, and the importance of this project in attracting further funding from (e.g.) the Sloan Foundation and T.\ Rowe Price, I am quite confident that we can produce a thorough report in the short time frame.
