       \documentclass[12pt,pdftex,letterpaper]{article}
            \usepackage{setspace}
            \usepackage[dvips,]{graphicx} %draft option suppresses graphics dvi display
%            \usepackage{lscape}
%            \usepackage{latexsym}
%            \usepackage{endnotes}
%            \usepackage{epsfig}
            \usepackage{amsmath}
%           \singlespace
            \setlength{\textwidth}{6.5in}
            \setlength{\textheight}{9in}
            \addtolength{\topmargin}{-\topmargin} 
            \setlength{\oddsidemargin}{0in}
            \setlength{\evensidemargin}{0in}
            \addtolength{\headsep}{-\headsep}
            \addtolength{\topskip}{-\topskip}
            \addtolength{\headheight}{-\headheight}
            \setcounter{secnumdepth}{2}
%            \renewcommand{\thesection}{\arabic{section}}
            % \renewcommand{\footnote}{\endnote}
            \newtheorem{proposition}{Proposition}
            \newtheorem{definition}{Definition}
            \newtheorem{lemma}{lemma}
            \newtheorem{corollary}{Corollary}
            \newtheorem{assumption}{Assumption}
            \newcommand{\Prob}{\operatorname{Prob}}
            \clubpenalty 5000
            \widowpenalty 5000
            \renewcommand{\baselinestretch}{1.1}
            \usepackage{amsmath}
            \usepackage{amsthm}
            \usepackage{amsfonts}
            \usepackage{amssymb}
            \usepackage{bbm}
            \newcommand{\N}{\mathbb{N}}
			\newcommand{\R}{\mathbb{R}}
			\newcommand{\E}{\mathbb{E}}
			\newcommand{\der}[2]{\frac{\text{d}#1}{\text{d}#2}}
			\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
			
			\setlength{\parindent}{0pt}
			\setlength{\parskip}{1em}
						
\begin{document}
	\pagestyle{empty}
	
Jim:

Jim:

"consumption Carroll" here!

Our previous email chain was the result of your contacting me to see if we needed funding for the [Econ-ARK](https://econ-ark.org) project.  At that time, we'd recently received a generous grant from the Sloan Foundation-- I think my response was along the lines of  "we need people, not money!''  Since then, we have found (some of) the right people and have made great progress.   

Having completed most of what we originally envisioned, we now have ambitious ideas for our next steps. (I'm attaching a brief letter of interest describing our proposed work, and its virtues.)  In short, Econ-ARK wants to create a modeling framework (like, a much more powerful version of DYNARE) that encompasses the wide array of features that can be included in either micro or macro dynamic structural models -- basically, we want to create a way to represent any Bellman problem.  The idea is to design a specification (that is both human- and machine-readable) for describing the mathematical content of models, as well as the numeric methods used to solve them.

I'm reaching out now because, if you're amenable, this is the right time to get help from Schmidt Futures, both with respect to funding and personnel. You mentioned that the organization has several Googlers emeriti, and their technical expertise could be tremendously valuable to us.  

It would be great if we could set up a meeting with you to discuss all this (or, someone else if that would be more appropriate).

Thank you very much, and I hope you are well.

~~~~~---CDC


\end{document}